Doublet splitting and linear amplitude characteristics of fusion alpha-particle driven ion cyclotron emission

### 2022-02-05 manuscript.tex
minor updates to manuscript

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 (ibid). The electrons and deuterons, both modelled as
Maxwellians, have temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds to a deuterium cyclotron frequency of $17MHz$.
The ring-beam distribution function, Eq. \ref{eq:alphaf}, is a separable function of $v_\parallel$, i.e. the beam given by Eq. 1, and $v_\perp$, i.e. the ring given by Eq. 2. The component parallel (perpendicular) to the magnetic field is a drifting Maxwellian with thermal speed $v_{\parallel, th}$ ($v_{\perp, th}$) and drift velocity $u_\parallel$ ($u_\perp$). The thermal spread of the alpha particles' distribution function $v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The concentration ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data in the $(k_\parallel,\omega)$ plane where shading instead indicates growth rates. The simplicity of the consecutive unstable cyclotron harmonics visible in panel (a) betrays the complexity of the structure in the $(k_\parallel, \omega)$ plane. Integration over $k_\parallel$ projects the solutions onto the frequency axis, which loses the structure and produces narrow harmonics. Fig. \ref{fig:2D_imag_zeropitch} plots shows the same data as Fig. \ref{fig:2D_F12_zeropitch} with damped and normal modes in addition, except shading indicates the growth rates in $(k_\parallel, k_\perp)$ space. Overlaid contours indicate the real frequency in units of ion cyclotron harmonics. Only narrow regions around the cyclotron harmonics are unstable. Overlaid straight traces indicate propagation angle. No Doppler shift is present due to the lack of parallel drift in the alpha-particle ring-beam. The growth rate increases approximately linearly with cyclotron harmonic number, which is consistent with the literature\cite{Dendy1994}. Together these figures serves to introduce a 2D view (growth rates in $(k_\parallel, k_\perp)$ space) of traditionally 1D data (growth rates against frequency). This will become important when a non-zero $u_\parallel$ is taken into account next.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig1.pdf}
    \caption{(colour online) Unstable fast Alfv{\'e}n wave solutions to the
    linear Maxwell-Vlasov system of equations, Eq. \ref{eq:dettensor}, for
    thermal electrons, deuterons and a ring distribution of energetic
    alpha-particles i.e. when there is zero parallel drift, $u_\parallel=0$ in
    Eq. 1. (a) Upper panel: growth rate as a function of frequency where shading
    indicates the parallel wavenumber in units of $\Omega_i/V_A$ of the complex frequency solution. Consecutive cyclotron harmonics are increasingly unstable and are narrow due to the lack of Doppler shift associated with parallel drifts. (b) Lower panel: the same solutions as in panel (a) in the $(k_\parallel,\omega)$ plane where shading indicates growth rate. The complex structure of the growth rates that exists in wavenumber space is projected into sharp harmonics on the frequency axis.}
    \label{fig:2D_F12_zeropitch}
\end{figure}

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift, $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$, and (b) plots the solutions
    in the $k_\parallel, \omega$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks because both
    the growth rate and real frequency vary with wavenumber. Compare with
    Fig. 2 of Cottrel et al 1993.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

@@

These calculations capture the finite perpendicular thermal spread of the alpha
particle distribution function, the integral of which has no known analytically
tractable form. The computational capabilities of the software,
PlasmaDispersionRelations.jl, used to generate the results further enable ICE as
a diagnostic for fusion plasmas since the code is capable of: calculating the
dispersion relation of plasmas consisting of an arbitrary number of species; and
of arbitrary distribution functions, although those with parallel Maxwellians,
like those presented here, are faster. It has been shown that it is possible to
recreate signals containing lower harmonics of substantial amplitudes by
invoking 2D linear physics. The presence of ICE signals at lower harmonics in
PIC simulations has been attributed to nonlinear interactions between pairs of
more powerful higher harmonics\cite{Carbajal2014}. This work suggests that
observations of lower ICE harmonics may arise either due to the effects of 2D
linear physics or to nonlinear 1D physics\cite{Chapman2018}, or a combination.
There are several important features missing from these calculations. Among them
are the effect the spatial inhomogeniety of the magnetic field and background
plasma, and the quantisation associatied with toroidal periodicity. These may
create an eigenmode structure into which the MCI must fit.
Cottrell et al. \cite{Cottrell1993} does discuss the implications of propagation angles on ICE signals and indicate that lower harmonics must propagate increasingly close to exactly perpendicular. These effects remain the subject of further work to incorporate them in linear and nonlinear studies of ICE.

@@

It is a pleasure to thank R O Dendy and B Chapman-Oplopoiou useful discussions.

>>>>

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 (ibid). The electrons and deuterons, both modelled as
Maxwellians, have temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds a major radius of $4m$.
The ring-beam distribution function, Eq. \ref{eq:alphaf}, is a separable function of $v_\parallel$, i.e. the beam given by Eq. 1, and $v_\perp$, i.e. the ring given by Eq. 2. The component parallel (perpendicular) to the magnetic field is a drifting Maxwellian with thermal speed $v_{\parallel, th}$ ($v_{\perp, th}$) and drift velocity $u_\parallel$ ($u_\perp$). The thermal spread of the alpha particles' distribution function $v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The concentration ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rates. The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$ plane. Integration over $k_\parallel$ projects the solutions onto the frequency axis, which loses the structure and produces narrow harmonics. Fig. \ref{fig:2D_imag_zeropitch} plots shows the same data as Fig. \ref{fig:2D_F12_zeropitch} with damped and normal modes in addition, except shading indicates the growth rates in $(k_\parallel, k_\perp)$ space. Overlaid contours indicate the real frequency in units of ion cyclotron harmonics. Only narrow regions around the cyclotron harmonics are unstable. Overlaid straight traces indicate propagation angle. No Doppler shift is present due to the lack of parallel drift in the alpha-particle ring-beam. The growth rate increases approximately linearly with cyclotron harmonic number, which is consistent with the literature\cite{Dendy1994}. Together these figures serves to introduce a 2D view (growth rates in $(k_\parallel, k_\perp)$ space) of traditionally 1D data (growth rates against frequency). This will become important when a non-zero $u_\parallel$ is taken into account next.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig1.pdf}
    \caption{(colour online) Unstable fast Alfv{\'e}n wave solutions to the
    linear Maxwell-Vlasov system of equations, Eq. \ref{eq:dettensor}, for
    thermal electrons, deuterons and a ring distribution of energetic
    alpha-particles i.e. when there is zero parallel drift, $u_\parallel=0$ in
    Eq. 1. (a) Upper panel: growth rate as a function of frequency where shading
    indicates the parallel wavenumber in units of $\Omega_i/V_A$ of the complex
    frequency solution. Consecutive cyclotron harmonics are increasingly
    unstable and are narrow due to the lack of Doppler shift associated with
    parallel drifts. (b) Lower panel: the same solutions as in panel (a) in the
    $(\omega,k_\parallel)$ plane where shading indicates growth rate. The complex structure of the growth rates that exists in wavenumber space is projected into sharp harmonics on the frequency axis.}
    \label{fig:2D_F12_zeropitch}
\end{figure}

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift, $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$, and (b) plots the solutions
    in the $(\omega,k_\parallel)$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks because both
    the growth rate and real frequency vary with wavenumber. Compare with
    Fig. 2 of Cottrel et al 1993.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

@@

These calculations capture the finite perpendicular thermal spread of the alpha
particle distribution function, the integral of which has no known analytically
tractable form. The computational capabilities of the software,
PlasmaDispersionRelations.jl, used to generate the results further enable ICE as
a diagnostic for fusion plasmas since the code is capable of: calculating the
dispersion relation of plasmas consisting of an arbitrary number of species; and
of arbitrary distribution functions, although those with parallel Maxwellians,
like those presented here, are faster. It has been shown that it is possible to
recreate signals containing lower harmonics of substantial amplitudes by
invoking 2D linear physics. The presence of ICE signals at lower harmonics in
PIC simulations has been attributed to nonlinear interactions between pairs of
more powerful higher harmonics\cite{Carbajal2014}. This work suggests that
observations of lower ICE harmonics may arise either due to the effects of 2D
linear physics or to nonlinear 1D physics\cite{Chapman2018}, or a combination.
There are several important features missing from these calculations. Among them
are the effect the spatial inhomogeniety of the magnetic field and background
plasma, and the quantisation associatied with toroidal periodicity. These
effects may create an eigenmode structure, not accounted for herein,
associated with the kinetic MCI.
Cottrell et al. \cite{Cottrell1993} does discuss the implications of propagation angles on ICE signals and indicate that lower harmonics must propagate increasingly close to exactly perpendicular. These effects remain the subject of further work to incorporate them in linear and nonlinear studies of ICE.

@@

It is a pleasure to thank R O Dendy and B Chapman-Oplopoiou for useful discussions.
